Accessing and Analyzing Linguistic Field Data

### 2006-10-31 slides/data.tex
further polishing
svn/trunk@3665

\begin{frame}[fragile]
\frametitle{Example: Microsoft Word}
\begin{itemize}
\item sleep \verb|[sli:p]| \textbf{vi} \textit{condition of body and mind...}
\item We can enter this in MSWord, then "Save as Web Page"
\item resulting HTML:

@@

\begin{verbatim}
>>> from nltk_lite.corpora import toolbox
>>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
\end{verbatim}
\end{frame}

@@

\begin{itemize}
\item lexicon is a series of \texttt{record} objects
\item each contains field objects, such as \texttt{lx} and \texttt{ps}
\item address all the lexemes: \texttt{record/lx}
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{Adding New Fields (cont)}
\small
\begin{verbatim}
>>> toolbox.to_sfm_string(lexicon[50])
\lx kaeviro
\cv CVVCVCV
\ps V.A
\ge lift off
\ge take off
\gp go antap
\nt used to describe action of plane
\dt 12/Feb/2005
\ex Pita kaeviroroe kepa kekesia oa vuripierevo kiuvu.
\xp Pita i go antap na lukim haus win i bagarapim.
\xe Peter went to look at the house that the wind destroyed.
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> html = "<table>\n"
>>> for entry in lexicon[70:80]:
...     lx = entry.findtext('lx')
...     ps = entry.findtext('ps')
...     ge = entry.findtext('ge')
...     html += "  <tr><td>%s</td><td>%s</td><td>%s</td></tr>\n" % (lx, ps, ge)
>>> html += "</table>"
>>> print html
\end{verbatim}

@@

\scriptsize
\begin{verbatim}
>>> import sys
>>> from nltk_lite.etree.ElementTree import ElementTree
>>> tree = ElementTree(lexicon[3])
>>> tree.write(sys.stdout)
<record>
  <lx>kaakaaro</lx>
  <ps>N.N</ps>
  <ge>mixtures</ge>
  <gp>???</gp>
  <eng>mixtures</eng>
  <eng>charm used to keep married men and women youthful and attractive</eng>
  <cmt>Check vowel length. Is it kaakaaro or kaakaro?</cmt>
  <dt>14/May/2005</dt>
  <ex>Kaakaroto ira purapaiveira aue iava opita, voeao-pa airepa oraouirara, ra va aiopaive.</ex>
  <xp>Kokonas ol i save wokim long ol kain samting bilong ol nupela marit, bai ol i ken kaikai.</xp>
  <xe>Mixtures are made from coconut, ???.</xe>
</record>
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> for lex in lexgloss:
...     if lex+lex in lexgloss:
...         print "%s (%s); %s (%s)" % (lex, lexgloss[lex], lex+lex, lexgloss[lex+lex])
\end{verbatim}

\tiny
\begin{verbatim}
kuvu (fill.up); kuvukuvu (stamp the ground)
kitu (save); kitukitu (scrub clothes)
kopa (ingest); kopakopa (gulp.down)
kasi (burn); kasikasi (angry)
koi (high pitched sound); koikoi (groan with pain)
kee (chip); keekee (shattered)
kauo (jump); kauokauo (jump up and down)
kea (deceived); keakea (lie)
kove (drop); kovekove (drip repeatedly)
kape (unable to meet); kapekape (grip with arms not meeting)
kapo (fasten.cover.strip); kapokapo (fasten.cover.strips)
koa (skin); koakoa (remove the skin)
kipu (paint); kipukipu (rub.on)
koe (spoon out a solid); koekoe (spoon out)
kovo (work); kovokovo (surround)
kiru (have sore near mouth); kirukiru (crisp)
kotu (bite); kotukotu (grind teeth together)
kavo (collect); kavokavo (work black magic)
kuri (scrape); kurikuri (scratch repeatedly)
karu (unhook); karukaru (open)
kare (return); karekare (return)
kari (break); karikari (shred)
kiro (write); kirokiro (write)
kae (carry); kaekae (tempt)
koru (make return); korukoru (obstruct)
ku (finished with); kuku (spoonfeed)
kosi (exit); kosikosi (exit)
\end{verbatim}
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Complex Search Criteria}
\begin{itemize}
\item for phonological description, identify segments, alternations,
  syllable canon...
\item what syllable types occur in lexemes (MSC, conspiracies)?
\scriptsize

\begin{verbatim}
>>> from nltk_lite.tokenize import regexp
>>> from nltk_lite.probability import FreqDist
>>> fd = FreqDist()
>>> lexemes = [lexeme.text.lower() for lexeme in lexicon.findall('record/lx')]
>>> for lex in lexemes:
...     for syl in regexp(lex, pattern=r'[^aeiou][aeiou]'):
...         fd.inc(syl)
\end{verbatim}
\end{itemize}
\end{frame}

@@

{\scriptsize
\begin{verbatim}
>>> for vowel in 'aeiou':
...     for cons in 'ptkvsr':
...          print '%s%s:%4d ' % (cons, vowel, fd.count(cons+vowel)),
...     print
pa:  84  ta:  43  ka: 414  va:  87  sa:   0  ra: 185 
pe:  32  te:   8  ke: 139  ve:  25  se:   1  re:  62 
pi:  97  ti:   0  ki:  88  vi:  96  si:  95  ri:  83 
po:  31  to: 140  ko: 403  vo:  42  so:   3  ro:  86 
pu:  49  tu:  35  ku: 169  vu:  44  su:   1  ru:  72 
\end{verbatim}}
\item NB \texttt{t} and \texttt{s} columns
\item \texttt{ti} not attested, while \texttt{si} is frequent: palatalization?
\item which lexeme contains \texttt{su}?  \textit{kasuari}
\end{itemize}
\end{frame}

@@

\begin{verbatim}
>>> sum(len(entry) for entry in lexicon) / len(lexicon)
10
\end{verbatim}
\end{frame}

@@

@@

\begin{frame}[fragile]
\frametitle{Looking at Timestamps}
\scriptsize
\begin{verbatim}
>>> fd = FreqDist()
>>> from string import split
>>> for entry in shoebox.dictionary('rotokas'):
...     if 'dt' in entry:
...         (day, month, year) = split(entry['dt'], '/')
...         fd.inc((month, year))
>>> for time in fd.sorted_samples():
...     print time[0], '/', time[1], ':', fd.count(time)
Feb / 2005 : 307
Dec / 2004 : 151
Jan / 2005 : 123
Feb / 2004 : 64
Sep / 2004 : 49
May / 2005 : 46
Mar / 2005 : 37
Apr / 2005 : 29
Jul / 2004 : 14
Nov / 2004 : 5
Oct / 2004 : 5
Aug / 2004 : 4
May / 2003 : 2
Jan / 2004 : 1
May / 2004 : 1
\end{verbatim}
\end{frame}

>>>>

\begin{frame}[fragile]
\frametitle{Example: Microsoft Word}
\small
\begin{itemize}
\item sleep \verb|[sli:p]| \textbf{vi} \textit{condition of body and mind...}
\item We can enter this in MSWord, then "Save as Web Page"
\item resulting HTML:

@@

\small
\begin{verbatim}
>>> from nltk_lite.corpora import toolbox
>>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
\end{verbatim}
\end{frame}

@@

\small
\begin{itemize}
\item lexicon is a series of \texttt{record} objects
\item each contains field objects, such as \texttt{lx} and \texttt{ps}
\item address all the lexemes: \texttt{record/lx}
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{Adding New Fields (cont)}
\scriptsize
\begin{verbatim}
>>> toolbox.to_sfm_string(lexicon[50])
\lx kaeviro
\cv CVVCVCV
\ps V.A
\ge lift off
\ge take off
\gp go antap
\nt used to describe action of plane
\dt 12/Feb/2005
\ex Pita kaeviroroe kepa kekesia oa vuripierevo kiuvu.
\xp Pita i go antap na lukim haus win i bagarapim.
\xe Peter went to look at the house that the wind destroyed.
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> html = "<table>\n"
>>> for entry in lexicon[70:80]:
...     lx = entry.findtext('lx')
...     ps = entry.findtext('ps')
...     ge = entry.findtext('ge')
...     html += "  <tr><td>%s</td><td>%s</td><td>%s</td></tr>\n" %
...       (lx, ps, ge)
>>> html += "</table>"
>>> print html
\end{verbatim}

@@

\tiny
\begin{verbatim}
>>> import sys
>>> from nltk_lite.etree.ElementTree import ElementTree
>>> tree = ElementTree(lexicon[3])
>>> tree.write(sys.stdout)
<record>
  <lx>kaakaaro</lx>
  <ps>N.N</ps>
  <ge>mixtures</ge>
  <gp>???</gp>
  <eng>mixtures</eng>
  <eng>charm used to keep married men and women youthful and attractive</eng>
  <cmt>Check vowel length. Is it kaakaaro or kaakaro?</cmt>
  <dt>14/May/2005</dt>
  <ex>Kaakaroto ira purapaiveira aue iava opita, voeao-pa airepa oraouirara, ra va aiopaive.</ex>
  <xp>Kokonas ol i save wokim long ol kain samting bilong ol nupela marit, bai ol i ken kaikai.</xp>
  <xe>Mixtures are made from coconut, ???.</xe>
</record>
\end{verbatim}
\end{frame}

@@

\begin{verbatim}
>>> for lex in lexgloss:
...     if lex+lex in lexgloss:
...         print "%s (%s); %s (%s)" % (lex, lexgloss[lex], lex+lex, lexgloss[lex+lex])
\end{verbatim}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Reduplication (cont)}
\scriptsize
\begin{verbatim}
kuvu (fill.up); kuvukuvu (stamp the ground)
kitu (save); kitukitu (scrub clothes)
kopa (ingest); kopakopa (gulp.down)
kasi (burn); kasikasi (angry)
koi (high pitched sound); koikoi (groan with pain)
kee (chip); keekee (shattered)
kauo (jump); kauokauo (jump up and down)
kea (deceived); keakea (lie)
kove (drop); kovekove (drip repeatedly)
kape (unable to meet); kapekape (grip with arms not meeting)
kapo (fasten.cover.strip); kapokapo (fasten.cover.strips)
koa (skin); koakoa (remove the skin)
kipu (paint); kipukipu (rub.on)
koe (spoon out a solid); koekoe (spoon out)
kovo (work); kovokovo (surround)
kiru (have sore near mouth); kirukiru (crisp)
kotu (bite); kotukotu (grind teeth together)
kavo (collect); kavokavo (work black magic)
...
\end{verbatim}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Complex Search Criteria}
\begin{itemize}
\item for phonological description, identify segments, alternations,
  syllable canon...
\item what syllable types occur in lexemes (MSC, conspiracies)?
\end{itemize}

\scriptsize
\begin{verbatim}
>>> from nltk_lite.tokenize import regexp
>>> from nltk_lite.probability import FreqDist
>>> fd = FreqDist()
>>> lexemes = [lexeme.text.lower() for lexeme in lexicon.findall('record/lx')]
>>> for lex in lexemes:
...     for syl in regexp(lex, pattern=r'[^aeiou][aeiou]'):
...         fd.inc(syl)
\end{verbatim}
\end{frame}

@@

\scriptsize
\begin{verbatim}
>>> for vowel in 'aeiou':
...     for cons in 'ptkvsr':
...          print '%s%s:%4d ' %
...            (cons, vowel, fd.count(cons+vowel)),
...     print
pa:  84  ta:  43  ka: 414  va:  87  sa:   0  ra: 185 
pe:  32  te:   8  ke: 139  ve:  25  se:   1  re:  62 
pi:  97  ti:   0  ki:  88  vi:  96  si:  95  ri:  83 
po:  31  to: 140  ko: 403  vo:  42  so:   3  ro:  86 
pu:  49  tu:  35  ku: 169  vu:  44  su:   1  ru:  72 
\end{verbatim}

\small
\item NB \texttt{t} and \texttt{s} columns
\item \texttt{ti} not attested, while \texttt{si} is frequent: palatalization?
\item which lexeme contains \texttt{su}?  \textit{kasuari}
\end{itemize}
\end{frame}

@@

\small
\begin{verbatim}
>>> sum(len(entry) for entry in lexicon) / len(lexicon)
10
\end{verbatim}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Grammar}
\scriptsize

@@

\begin{frame}[fragile]
\frametitle{Looking at Timestamps}

\scriptsize
\begin{verbatim}
>>> fd = FreqDist()
>>> from string import split
>>> for entry in shoebox.dictionary('rotokas'):
...     if 'dt' in entry:
...         (day, month, year) = split(entry['dt'], '/')
...         fd.inc((month, year))
>>> for time in fd.sorted_samples():
...     print time[0], '/', time[1], ':', fd.count(time)
Feb / 2005 : 307
Dec / 2004 : 151
Jan / 2005 : 123
Feb / 2004 : 64
Sep / 2004 : 49
May / 2005 : 46
Mar / 2005 : 37
Apr / 2005 : 29
Jul / 2004 : 14
Nov / 2004 : 5
Oct / 2004 : 5
...
\end{verbatim}
\end{frame}
